\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3.2\baselineskip}

### 2017-03-25 paper/ms.tex
addressing comments from @jakevdp

% Projects:
\newcommand{\project}[1]{\textsf{#1}}
\newcommand{\kepler}{\project{Kepler}}
\newcommand{\tess}{\project{TESS}}
\newcommand{\celerite}{\project{celerite}}
\newcommand{\emcee}{\project{emcee}}

@@

In this paper, we present a method for computing a class of GP models that
scales linearly with the number of data points $\mathcal{O}(N)$ for one
dimensional data sets.
This method is a generalization of a method developed by
\citet{Ambikasaran:2015} that was, in turn, built on intuition from a twenty
year old paper \citep{Rybicki:1995}.
For this method to be applicable, the data must be one-dimensional and the
covariance function must have a specific form.
However, there is no further constraint on the data or the model.
In particular, the measurements don't need to be evenly spaced and the
uncertainties can be heteroscedastic (non-uniform).
This method is especially appealing compared to other similar methods~--~we
return to these below~--~because it is exact, flexible, simple, and fast.

@@

By rewriting the exponentials in \eq{celerite-kernel-complex} as sums of sine
and cosine functions, we can see the autocorrelation structure is defined by a
mixture of quasiperiodic oscillators
\begin{eqnarray}\eqlabel{celerite-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
For clarity, we refer to the argument within the sum as a ``\celerite\
term'' for the remainder of this paper.
The Fourier transform\footnote{Here and throughout we have defined the Fourier
transform of the function $f(t)$ as $F(\omega)={(2\,\pi)}^{-1/2}\,
\int_{-\infty}^\infty f(t)\,e^{i\,\omega\,t}\dd t$.} of this covariance
function is the power spectral density (PSD) of the process and it is given by
\begin{eqnarray}\eqlabel{celerite-psd}
S(\omega) = \sum_{j=1}^J \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+{({c_j}^2+{d_j}^2)}^2}\quad.
\end{eqnarray}
The physical interpretation of this model isn't immediately obvious and we
return to a more general discussion of the physical intuition in a moment
but we start by investigating some special cases.

@@

Finally, we note that the product of two terms of the form found inside the
sum in \eq{celerite-kernel} can also be re-written as a sum with updated
parameters
\begin{eqnarray}\eqlabel{product-rule}
k_j(\tau) \, k_k(\tau) =
    e^{-\tilde{c}\,\tau}\,[
        \tilde{a}_+\,\cos(\tilde{d}_+\,\tau) + \tilde{b}_+\,\sin(\tilde{d}_+\,\tau) +
        \tilde{a}_-\,\cos(\tilde{d}_-\,\tau) + \tilde{b}_-\,\sin(\tilde{d}_-\,\tau)
    ]
\end{eqnarray}
where
\begin{eqnarray}
    \tilde{a}_{\pm} &=& \frac{1}{2}\,(a_j\,a_k \pm b_j\,b_k) \\
    \tilde{b}_{\pm} &=& \frac{1}{2}\,(b_j\,a_k \mp a_j\,b_k) \\
    \tilde{c} &=& c_j + c_k \\
    \tilde{d}_{\pm} &=& d_j \mp d_k \quad.
\end{eqnarray}
Therefore, the method described in the following section can be used to
perform scalable inference on large datasets for any model, where the kernel
function is a sum or product of \celerite\ terms.

@@

In the case of the full \celerite\ covariance function
(\eqalt{celerite-kernel}), we introduce the following auxiliary variables in
analogy to the $u_n$ and $g_n$ that we introduced in the previous section
\begin{eqnarray}\eqlabel{full-system-first}
\phi_{n,j} &=& e^{-c_j\,\tau_{n+1,n}}\,\cos\left(d_j\,\tau_{n+1,n}\right)\\
\psi_{n,j} &=& -e^{-c_j\,\tau_{n+1,n}}\,\sin\left(d_j\,\tau_{n+1,n}\right)\\
g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,z_n + \psi_{n,j}\,h_{n-1,j}\\
h_{n,j} &=& \phi_{n,j}\,h_{n-1,j} - \psi_{n,j}\,z_n - \psi_{n,j}\,g_{n-1,j}\\
u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + a_j\,z_{n+1} + \psi_{n+1,j}\,v_{n+1,j}\\
v_{n,j} &=& \phi_{n+1,j}\,v_{n+1,j} - b_j\,z_{n+1} - \psi_{n+1,j}\,u_{n+1,j}
\end{eqnarray}
with the boundary conditions
\begin{eqnarray}
    g_{0,j} = 0 \quad, \quad
    h_{0,j} = 0 \quad, \quad
    u_{N,j} = 0 \quad, \quad\mathrm{and}\quad
    v_{N,j} = 0
\end{eqnarray}
for all $j$.
Using these variables and some algebra, we find that the following expression
\begin{eqnarray}\eqlabel{full-system-last}
\sum_{j=1}^J \left[a_j\,g_{n,j}+b_j\,h_{n,j}\right]
+ \left[{\sigma_n}^2+\sum_{j=1}^J a_j\right]
+ \sum_{j=1}^J \left[\phi_{n,j}\,u_{n,j}+\psi_{n,j}\,v_{n,j}\right]
    &=& r_{\bvec{\theta},n}
\end{eqnarray}
is equivalent to the target matrix equation
\begin{eqnarray}
K\,\bvec{z} &=& \bvec{r}_\bvec{\theta}
\end{eqnarray}
if $r_{\bvec{\theta},n}$ is the $n$-th element of the residual vector
$\bvec{r}_\bvec{\theta}$ defined in \sect{gps}.
\eq{full-system-first} through \eq{full-system-last} define a banded matrix
equation in the ``extended'' space and, as before, this can be used to solve
for $K^{-1}\,\bvec{r}_\bvec{\theta}$ and $\det K$ in $\mathcal{O}(N)$
operations.
\Figure{matrix} shows a pictorial representation of the sparsity pattern of
the extended matrix $K_\mathrm{ext}$.
Given this definition of $K_\mathrm{ext}$, the corresponding extended vectors
$\bvec{z}_\mathrm{ext}$ and $\bvec{r}_\mathrm{ext}$ are defined schematically
as
\begin{eqnarray}
{\bvec{z}_\mathrm{ext}} ^\T =
\left(\begin{array}{cccccccccc}
    z_1 & u_{1,j} & v_{1,j} & g_{1,j} & h_{1,j} & z_2 & u_{2,j} & \cdots &
    h_{N-1,j} & z_N
\end{array}\right)
\end{eqnarray}
and
\begin{eqnarray}
{\bvec{r}_\mathrm{ext}} ^\T =
\left(\begin{array}{cccccccccc}
    r_{\bvec{\theta},1} & 0 & 0 & 0 & 0 & r_{\bvec{\theta},2} &
    0 & \cdots & 0 & r_{\bvec{\theta},N}
\end{array}\right) \quad.
\end{eqnarray}
After constructing the extended matrix (using a compact storage format), the
extended matrix can be factorized using a LU-decomposition\footnote{Even
though $K_\mathrm{ext}$ is symmetric, it is not positive definite so a
Cholesky solver cannot be used for increased efficiency.} routine optimized
for band or sparse matrices.
This decomposition can then be used to compute the determinant of $K$, solve
$K^{-1}\,\bvec{r}_\bvec{\theta}$, and subsequently calculate the marginalized
likelihood in \eq{gp-likelihood}.

@@

We now turn to a discussion of \celerite\ as a model of astrophysical
variability.
A common concern in the context of GP modeling in astrophysics is the lack of
physical motivation for the choice of kernel functions.
Kernels are often chosen simply because they are popular with little
consideration of the impact of this decision.
While this isn't necessarily a problem for the inferred results, in this
section we discuss an exact physical interpretation of the \celerite\ kernel
that will be applicable to many astrophysical systems but especially the
time-variability of stars.

@@

To relate this physical picture to \celerite, we consider the dynamics of a
stochastically-driven damped simple harmonic oscillator.
The differential equation for this system is
\begin{equation}
    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
    + \omega_0^2\right]\, y(t) = \epsilon(t)
\end{equation}
where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
force.
If $\epsilon(t)$ is white noise, the PSD of this process is
\citep{Anderson:1990}
\begin{equation}\eqlabel{sho-psd}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{equation}
where $S_0$ is proportional to the power at $\omega = \omega_0$, $S(\omega_0)
= \sqrt{2/\pi}\,S_0\,Q^2$.
The power spectrum in \eq{sho-psd} matches \eq{celerite-psd} if
\begin{eqnarray}\eqlabel{sho-complex}
a_j &=& S_0\,\omega_0\,Q \\
b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
c_j &=& \frac{\omega_0}{2\,Q}\\
d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
\end{eqnarray}
for $Q \ge \frac{1}{2}$.
For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of \celerite\
terms with parameters
\begin{eqnarray}\eqlabel{sho-real}
a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
        \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
b_{j\pm} &=& 0 \nonumber\\
    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right]
    \nonumber\\
d_{j\pm} &=& 0 \quad. \nonumber
\end{eqnarray}

@@

{\item Substituting $Q = 1/2$, \eq{sho-psd} becomes
\begin{eqnarray}
S(\omega) =
    \sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2+1\right]^2}
\end{eqnarray}
with the corresponding covariance function (using
\eqalt{celerite-kernel} and \eqalt{sho-real})
\begin{eqnarray}\eqlabel{approx-matern}
k(\tau) &=& \lim_{f \to 0}\,
    \frac{1}{2}\,S_0\,\omega_0\,
    \left[\left(1+1/f\right)\,e^{-\omega_0\,(1-f)\,\tau} +
          \left(1-1/f\right)\,e^{-\omega_0\,(1+f)\,\tau}
    \right] \\
&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau]
\end{eqnarray}
or equivalently (using \eqalt{celerite-kernel} and \eqalt{sho-complex})
\begin{eqnarray}\eqlabel{approx-matern2}
k(\tau) &=& \lim_{f \to 0}\,
    S_0\,\omega_0\,e^{-\omega_0\,\tau}\,
    \left[\cos(f\,\tau) + \frac{\omega_0}{f}\,\sin(f\,\tau)\right] \\
&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau] \quad.
\end{eqnarray}
This covariance function is also known as the Mat\'ern-3/2 function
\citep{Rasmussen:2006}.
This suggests that the Mat\'ern-3/2 covariance can be approximated using
the \celerite\ framework with a small value of $f$ in \eq{approx-matern2} but we
caution that this might lead to numerical issues for the solver.
}

@@

To demonstrate the application of \celerite, we start by inferring
posterior constraints on the parameters of a GP model applied to several
simulated datasets with known properties.
In the following section, we expand on these examples by applying
\celerite\ to real datasets.
In the first example, we demonstrate that \celerite\ can be used to infer
the power spectrum of a process when the data are generated from a \celerite\
model.
In the second example, we demonstrate that \celerite\ can be used as an
effective model even if the true process cannot be represented in the space of
allowed models.
This is an interesting example because, when analyzing real data, we rarely
have any fundamental reason to believe that the data were generated by a GP
model with a specific kernel.
Even in these cases, GPs can be useful effective models and \celerite\
provides computational advantages over other GP methods.

@@

In this first example, we simulate a dataset using a known \celerite\
process and fit it with \celerite\ to demonstrate that valid inferences can be
made in this idealized case.
The simulated dataset is shown in the left panel of \Figure{simulated-correct}
and it was generated using a SHO kernel (\eqalt{sho-kernel}) with parameters
$S_0 = 1$, $\omega_0 = e^2$, and $Q = e^2$.
The true PSD is shown as a dashed line in the right panel of
\Figure{simulated-correct}.
We applied log-uniform priors to all of the parameters and used \emcee\
\citep{Foreman-Mackey:2013} to sample the joint posterior density and computed
the marginalized posterior inference of the PSD.
This inference is shown in the right panel of \Figure{simulated-correct} as a
blue contour indicating 68\% of the posterior mass.
It is clear from this figure that, as expected, the inference correctly
reproduces the true PSD.

@@

For this example, we simulate a dataset using a known GP model with a kernel
outside of the support of a \celerite\ process.
This means the true autocorrelation of the process can never be correctly
represented by the model that we're using to fit but we use this example
to demonstrate that, at least in this case, valid inferences can still be made
about the physical parameters of the model.

For this example the data are simulated from a quasiperiodic GP with the
kernel
\begin{eqnarray}\eqlabel{sim-wrong-true}
k_\mathrm{true} (\tau) = \alpha\,
    \exp\left(-\frac{\tau^2}{2\,\lambda^2}\right)\,
    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{true}}\right)
\end{eqnarray}
where $P_\mathrm{true}$ is the fundamental period of the process.
This autocorrelation structure corresponds to the power spectrum
\begin{eqnarray}
S_\mathrm{true} (\omega) = \frac{\lambda\,\alpha}{2}\,\left[
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega-
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right) +
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega+
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right)
\right]
\end{eqnarray}
which, for large values of $\omega$, falls off exponentially.
When compared to \eq{celerite-psd}~--~which, for large $\omega$, goes as
$\omega^{-4}$ at most~--~it is clear that a \celerite\ model can never
perfectly reproduce the structure of this process.
That being said, we demonstrate that rigorous inferences can be made
about $P_\mathrm{true}$ even with an effective model.
The left panel of \Figure{simulated-wrong} shows the simulated dataset.
We then fit this simulated data using the product of two SHO terms
(\eqalt{sho-kernel}) where one of the terms has $S_0 = 1$ and $Q =
1/\sqrt{2}$ and the other has $\omega_0 = 2\,\pi/P$.
We note that using \eq{product-rule}, the product of two \celerite\ terms can
also be expressed using \celerite.
We apply log-uniform priors on all the parameters and use \emcee\ to sample
the joint posterior probability for all of the parameters.
The inferred distribution for the parameter $P$ is shown in the right panel of
\Figure{simulated-wrong} and compared to the true period $P_\mathrm{true}$ and
the inferences made using the correct model (\eqalt{sim-wrong-true}).
The inference made using this effective \celerite\ model are indistinguishable
from the inferences made using the correct model but substantially less
computation time is required for the \celerite\ inference.

@@

Gaussian Process models have been fruitfully applied to many problems in
astronomical data analysis but the fact that the computational cost scales as
the cube of the number of data points has limited their use to relatively
small datasets.
With the linear scaling of \celerite\, we envision the application of Gaussian
processes may be expanded to much larger datasets.
Despite the restrictive form of the \celerite\ kernel, with a sufficient
number of components it is flexible enough to describe a wide range of
astrophysical variability.
In fact, the relation of the \celerite\ kernel to the damped, stochastically-driven harmonic
oscillator matches simple models of astrophysical variability, and makes the
parameterization interpretable in terms of resonant frequency, amplitude, and
quality factor.
% Siva: Removed major, since $J$ is most often small as stated in the second half of the sentence
A drawback of this method is its quadratic scaling with the number of
terms $J$ but, in many cases, small values of $J$ are sufficient.

>>>>

% Projects:
\newcommand{\project}[1]{\textsf{#1}}
\newcommand{\kepler}{\project{Kepler}}
\newcommand{\tess}{\project{TESS}}
\newcommand{\celerite}{\project{celerite}}
\newcommand{\celeriteterm}{\emph{celerite}}
\newcommand{\emcee}{\project{emcee}}

@@

In this paper, we present a method for computing a class of GP models that
scales linearly with the number of data points $\mathcal{O}(N)$ for one
dimensional data sets.
This method is a generalization of a method developed by
\citet{Ambikasaran:2015} that was, in turn, built on intuition from a twenty
year old paper \citep{Rybicki:1995}.
For this method to be applicable, the data must be one-dimensional and the
covariance function must be represented by a mixture of exponentials.
However, there is no further constraint on the data or the model.
In particular, the measurements don't need to be evenly spaced and the
uncertainties can be heteroscedastic (non-uniform).

This method achieves linear scaling by exploiting structure in the covariance
matrix when it is generated by a mixture of exponentials.
More precisely, the matrix can be embedded in an equivalent higher-dimensional
system that can be solved efficiently using a sparse or banded solver.
As we will discuss in the following pages, this choice of kernel function is
not as restrictive as it first seems because it arises naturally in some
physical systems and it can be used as an effective model in many other cases.
This method is especially appealing compared to other similar methods~--~we
return to these below~--~because it is exact, flexible, simple, and fast.

@@

By rewriting the exponentials in \eq{celerite-kernel-complex} as sums of sine
and cosine functions, we can see the autocorrelation structure is defined by a
mixture of quasiperiodic oscillators
\begin{eqnarray}\eqlabel{celerite-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
For clarity, we refer to the argument within the sum as a ``\celeriteterm\
term'' for the remainder of this paper.
The Fourier transform\footnote{Here and throughout we have defined the Fourier
transform of the function $f(t)$ as $F(\omega)={(2\,\pi)}^{-1/2}\,
\int_{-\infty}^\infty f(t)\,e^{i\,\omega\,t}\dd t$.} of this covariance
function is the power spectral density (PSD) of the process and it is given by
\begin{eqnarray}\eqlabel{celerite-psd}
S(\omega) = \sum_{j=1}^J \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+{({c_j}^2+{d_j}^2)}^2}\quad.
\end{eqnarray}
The physical interpretation of this model isn't immediately obvious and we
return to a more general discussion of the physical intuition in a moment
but we start by investigating some special cases.

@@

Finally, we note that the product of two terms of the form found inside the
sum in \eq{celerite-kernel} can also be re-written as a sum with updated
parameters
\begin{eqnarray}\eqlabel{product-rule}
k_j(\tau) \, k_k(\tau) =
    e^{-\tilde{c}\,\tau}\,[
        \tilde{a}_+\,\cos(\tilde{d}_+\,\tau) + \tilde{b}_+\,\sin(\tilde{d}_+\,\tau) +
        \tilde{a}_-\,\cos(\tilde{d}_-\,\tau) + \tilde{b}_-\,\sin(\tilde{d}_-\,\tau)
    ]
\end{eqnarray}
where
\begin{eqnarray}
    \tilde{a}_{\pm} &=& \frac{1}{2}\,(a_j\,a_k \pm b_j\,b_k) \\
    \tilde{b}_{\pm} &=& \frac{1}{2}\,(b_j\,a_k \mp a_j\,b_k) \\
    \tilde{c} &=& c_j + c_k \\
    \tilde{d}_{\pm} &=& d_j \mp d_k \quad.
\end{eqnarray}
Therefore, the method described in the following section can be used to
perform scalable inference on large datasets for any model, where the kernel
function is a sum or product of \celeriteterm\ terms.

@@

In the case of the full \celeriteterm\ covariance function
(\eqalt{celerite-kernel}), we introduce the following auxiliary variables in
analogy to the $u_n$ and $g_n$ that we introduced in the previous section
\begin{eqnarray}\eqlabel{full-system-first}
\phi_{n,j} &=& e^{-c_j\,\tau_{n+1,n}}\,\cos\left(d_j\,\tau_{n+1,n}\right)\\
\psi_{n,j} &=& -e^{-c_j\,\tau_{n+1,n}}\,\sin\left(d_j\,\tau_{n+1,n}\right)\\
g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,z_n + \psi_{n,j}\,h_{n-1,j}\\
h_{n,j} &=& \phi_{n,j}\,h_{n-1,j} - \psi_{n,j}\,z_n - \psi_{n,j}\,g_{n-1,j}\\
u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + a_j\,z_{n+1} + \psi_{n+1,j}\,v_{n+1,j}\\
v_{n,j} &=& \phi_{n+1,j}\,v_{n+1,j} - b_j\,z_{n+1} - \psi_{n+1,j}\,u_{n+1,j}
\end{eqnarray}
with the boundary conditions
\begin{eqnarray}
    g_{0,j} = 0 \quad, \quad
    h_{0,j} = 0 \quad, \quad
    u_{N,j} = 0 \quad, \quad\mathrm{and}\quad
    v_{N,j} = 0
\end{eqnarray}
for all $j$.
Using these variables and some algebra, we find that the following expression
\begin{eqnarray}\eqlabel{full-system-last}
\sum_{j=1}^J \left[a_j\,g_{n,j}+b_j\,h_{n,j}\right]
+ \left[{\sigma_n}^2+\sum_{j=1}^J a_j\right]
+ \sum_{j=1}^J \left[\phi_{n,j}\,u_{n,j}+\psi_{n,j}\,v_{n,j}\right]
    &=& r_{\bvec{\theta},n}
\end{eqnarray}
is equivalent to the target matrix equation
\begin{eqnarray}
K\,\bvec{z} &=& \bvec{r}_\bvec{\theta}
\end{eqnarray}
if $r_{\bvec{\theta},n}$ is the $n$-th element of the residual vector
$\bvec{r}_\bvec{\theta}$ defined in \sect{gps}.
\eq{full-system-first} through \eq{full-system-last} define a banded matrix
equation in the ``extended'' space and, as before, this can be used to solve
for $K^{-1}\,\bvec{r}_\bvec{\theta}$ and $\det K$ in $\mathcal{O}(N)$
operations.
\Figure{matrix} shows a pictorial representation of the sparsity pattern of
the extended matrix $K_\mathrm{ext}$.
The extended matrix has the dimension $N_\mathrm{ext} = (4\,J+1)\,N-4\,J$,
with a bandwidth $W = 2\,J+2$ and $N_\mathrm{nz} = (20\,J+1)\,N-28\,J$
non-zero entries.
Given this definition of $K_\mathrm{ext}$, the corresponding extended vectors
$\bvec{z}_\mathrm{ext}$ and $\bvec{r}_\mathrm{ext}$ are defined schematically
as
\begin{eqnarray}
{\bvec{z}_\mathrm{ext}} ^\T =
\left(\begin{array}{cccccccccc}
    z_1 & u_{1,j} & v_{1,j} & g_{1,j} & h_{1,j} & z_2 & u_{2,j} & \cdots &
    h_{N-1,j} & z_N
\end{array}\right)
\end{eqnarray}
and
\begin{eqnarray}
{\bvec{r}_\mathrm{ext}} ^\T =
\left(\begin{array}{cccccccccc}
    r_{\bvec{\theta},1} & 0 & 0 & 0 & 0 & r_{\bvec{\theta},2} &
    0 & \cdots & 0 & r_{\bvec{\theta},N}
\end{array}\right) \quad.
\end{eqnarray}
After constructing the extended matrix (using a compact storage format), the
extended matrix can be factorized using a LU-decomposition\footnote{Even
though $K_\mathrm{ext}$ is symmetric, it is not positive definite so a
Cholesky solver cannot be used for increased efficiency.} routine optimized
for band or sparse matrices.
This decomposition can then be used to compute the determinant of $K$, solve
$K^{-1}\,\bvec{r}_\bvec{\theta}$, and subsequently calculate the marginalized
likelihood in \eq{gp-likelihood}.

@@

We now turn to a discussion of \celeriteterm\ as a model of astrophysical
variability.
A common concern in the context of GP modeling in astrophysics is the lack of
physical motivation for the choice of kernel functions.
Kernels are often chosen simply because they are popular with little
consideration of the impact of this decision.
While this isn't necessarily a problem for the inferred results, in this
section we discuss an exact physical interpretation of the \celeriteterm\ kernel
that will be applicable to many astrophysical systems but especially the
time-variability of stars.

@@

To relate this physical picture to \celeriteterm, we consider the dynamics of a
stochastically-driven damped simple harmonic oscillator.
The differential equation for this system is
\begin{equation}
    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
    + \omega_0^2\right]\, y(t) = \epsilon(t)
\end{equation}
where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
force.
If $\epsilon(t)$ is white noise, the PSD of this process is
\citep{Anderson:1990}
\begin{equation}\eqlabel{sho-psd}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{equation}
where $S_0$ is proportional to the power at $\omega = \omega_0$, $S(\omega_0)
= \sqrt{2/\pi}\,S_0\,Q^2$.
The power spectrum in \eq{sho-psd} matches \eq{celerite-psd} if
\begin{eqnarray}\eqlabel{sho-complex}
a_j &=& S_0\,\omega_0\,Q \\
b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
c_j &=& \frac{\omega_0}{2\,Q}\\
d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
\end{eqnarray}
for $Q \ge \frac{1}{2}$.
For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of
\celeriteterm\
terms with parameters
\begin{eqnarray}\eqlabel{sho-real}
a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
        \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
b_{j\pm} &=& 0 \nonumber\\
    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right]
    \nonumber\\
d_{j\pm} &=& 0 \quad. \nonumber
\end{eqnarray}

@@

{\item Substituting $Q = 1/2$, \eq{sho-psd} becomes
\begin{eqnarray}
S(\omega) =
    \sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2+1\right]^2}
\end{eqnarray}
with the corresponding covariance function (using
\eqalt{celerite-kernel} and \eqalt{sho-real})
\begin{eqnarray}\eqlabel{approx-matern}
k(\tau) &=& \lim_{f \to 0}\,
    \frac{1}{2}\,S_0\,\omega_0\,
    \left[\left(1+1/f\right)\,e^{-\omega_0\,(1-f)\,\tau} +
          \left(1-1/f\right)\,e^{-\omega_0\,(1+f)\,\tau}
    \right] \\
&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau]
\end{eqnarray}
or equivalently (using \eqalt{celerite-kernel} and \eqalt{sho-complex})
\begin{eqnarray}\eqlabel{approx-matern2}
k(\tau) &=& \lim_{f \to 0}\,
    S_0\,\omega_0\,e^{-\omega_0\,\tau}\,
    \left[\cos(f\,\tau) + \frac{\omega_0}{f}\,\sin(f\,\tau)\right] \\
&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau] \quad.
\end{eqnarray}
This covariance function is also known as the Mat\'ern-3/2 function
\citep{Rasmussen:2006}.
This suggests that the Mat\'ern-3/2 covariance can be approximated using
the \celeriteterm\ framework with a small value of $f$ in \eq{approx-matern2} but we
caution that this might lead to numerical issues for the solver.
}

@@

To demonstrate the application of \celerite, we start by inferring
posterior constraints on the parameters of a GP model applied to several
simulated datasets with known properties.
In the following section, we expand on these examples by applying
\celerite\ to real datasets.
In the first example, we demonstrate that \celerite\ can be used to infer
the power spectrum of a process when the data are generated from a
\celeriteterm\
model.
In the second example, we demonstrate that \celerite\ can be used as an
effective model even if the true process cannot be represented in the space of
allowed models.
This is an interesting example because, when analyzing real data, we rarely
have any fundamental reason to believe that the data were generated by a GP
model with a specific kernel.
Even in these cases, GPs can be useful effective models and \celerite\
provides computational advantages over other GP methods.

@@

In this first example, we simulate a dataset using a known \celeriteterm\
process and fit it with \celerite\ to demonstrate that valid inferences can be
made in this idealized case.
The simulated dataset is shown in the left panel of \Figure{simulated-correct}
and it was generated using a SHO kernel (\eqalt{sho-kernel}) with parameters
$S_0 = 1$, $\omega_0 = e^2$, and $Q = e^2$.
The true PSD is shown as a dashed line in the right panel of
\Figure{simulated-correct}.
We applied log-uniform priors to all of the parameters and used \emcee\
\citep{Foreman-Mackey:2013} to sample the joint posterior density and computed
the marginalized posterior inference of the PSD.
This inference is shown in the right panel of \Figure{simulated-correct} as a
blue contour indicating 68\% of the posterior mass.
It is clear from this figure that, as expected, the inference correctly
reproduces the true PSD.

@@

For this example, we simulate a dataset using a known GP model with a kernel
outside of the support of a \celeriteterm\ process.
This means the true autocorrelation of the process can never be correctly
represented by the model that we're using to fit but we use this example
to demonstrate that, at least in this case, valid inferences can still be made
about the physical parameters of the model.

For this example the data are simulated from a quasiperiodic GP with the
kernel
\begin{eqnarray}\eqlabel{sim-wrong-true}
k_\mathrm{true} (\tau) = \alpha\,
    \exp\left(-\frac{\tau^2}{2\,\lambda^2}\right)\,
    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{true}}\right)
\end{eqnarray}
where $P_\mathrm{true}$ is the fundamental period of the process.
This autocorrelation structure corresponds to the power spectrum
\begin{eqnarray}
S_\mathrm{true} (\omega) = \frac{\lambda\,\alpha}{2}\,\left[
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega-
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right) +
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega+
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right)
\right]
\end{eqnarray}
which, for large values of $\omega$, falls off exponentially.
When compared to \eq{celerite-psd}~--~which, for large $\omega$, goes as
$\omega^{-4}$ at most~--~it is clear that a \celeriteterm\ model can never
perfectly reproduce the structure of this process.
That being said, we demonstrate that rigorous inferences can be made
about $P_\mathrm{true}$ even with an effective model.
The left panel of \Figure{simulated-wrong} shows the simulated dataset.
We then fit this simulated data using the product of two SHO terms
(\eqalt{sho-kernel}) where one of the terms has $S_0 = 1$ and $Q =
1/\sqrt{2}$ and the other has $\omega_0 = 2\,\pi/P$.
We note that using \eq{product-rule}, the product of two \celeriteterm\ terms can
also be expressed using \celerite.
We apply log-uniform priors on all the parameters and use \emcee\ to sample
the joint posterior probability for all of the parameters.
The inferred distribution for the parameter $P$ is shown in the right panel of
\Figure{simulated-wrong} and compared to the true period $P_\mathrm{true}$ and
the inferences made using the correct model (\eqalt{sim-wrong-true}).
The inference made using this effective \celeriteterm\ model are indistinguishable
from the inferences made using the correct model but substantially less
computation time is required for the \celerite\ inference.

@@

Gaussian Process models have been fruitfully applied to many problems in
astronomical data analysis but the fact that the computational cost scales as
the cube of the number of data points has limited their use to relatively
small datasets.
With the linear scaling of \celerite\, we envision the application of Gaussian
processes may be expanded to much larger datasets.
Despite the restrictive form of the \celeriteterm\ kernel, with a sufficient
number of components it is flexible enough to describe a wide range of
astrophysical variability.
In fact, the relation of the \celeriteterm\ kernel to the damped, stochastically-driven harmonic
oscillator matches simple models of astrophysical variability, and makes the
parameterization interpretable in terms of resonant frequency, amplitude, and
quality factor.
% Siva: Removed major, since $J$ is most often small as stated in the second half of the sentence
A drawback of this method is its quadratic scaling with the number of
terms $J$ but, in many cases, small values of $J$ are sufficient.
